---
title: "Bio-assay experiment"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load the needed libraries into R 
```{r}
library(ggplot2)
library(StanHeaders)
library(rstan)

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

**1. Define the model and sample from posterior **

Note. There are 2 ways to do the model. One uses ready made functions inv_logit and binomial_logit (see Stan manual). The other 2 uses just definitions. This way is somewhat unstable and you need to adjust control parameters to get the model run so the first way is better
```{r}
model_bioassay = "
"
```

Load the data and put it into a list format
```{r}
bioassay <- read.table ("ex_bio-assay.dat", header=TRUE)
bioassay
N <- nrow(bioassay)
x <- bioassay$Dose
n <- bioassay$Nanim
y <- bioassay$Ndeath

data <- list ("N"=N, "x"=x, "n"=n, "y"=y)

## Define parameters and set initial values for them
## We are going to sample four chains so we need four starting points
## It is good practice to set them far apart from each others


## Run the Markov chain sampling with Stan:
post=stan(model_code=model_bioassay,data=data,warmup=500,iter=2000,init=inits,chains=4,thin=5)#,control=list(adapt_delta=0.9))

## check the convergence visually and by using PSRF by Gelman and Rubin (Rhat in Stan)

## visualize posterior (together and one at a time)

## Now parameter a contains a sample from the posterior p(a|y,x,n)
## and parameter b contains sample from the posterior p(b|y,x,n)
## We can now plot them in 2D and plot histograms of them

```

**2. visualize the posterior of theta as a function of dose level x**
```{r}
# See help from the linear regression example
# mean and appropriate posterior interval
# NOTE! 
# the posterior of theta at specific x is not Gaussian as 
# that of f in the linear regression example. Thus, use 
# quantile instead of std when calculating the 95% interval.
xp = seq(-0.8, 0.8, length=101)  # the evaluation points
mth = 0                       # initialize mean
q_lower = 0                      # initialize confidence interval
q_upper = 0                      # initialize 

for (i in 1:length(xp)) {

  }                        # new figure frame

```

**3. Calculate the covariance between a and b**

```{r}

```


**4. sample from the posterior distribution of LD50 and visualize it **


```{r}
## Now we will approximate the posterior of LD50 
## Sample from p(LD50 | x, y, n) and visualize that
##write the LD50 dose level as function of a and b
## see book page 77


```


**5. visualize the posterior distribution of theta with dose level (log g/ml) 0**


```{r}

```

**6. Posterior predictive distribution for new observation**

 Assume that we want to make new experiments to get better estimate for the LD50 statistics. What is the probability distribution for outcome of new experiments $p( \tilde{y} | \tilde{x},\tilde{n}, x, y, n)$?


```{r}


```

