---
title: "Exercise: censored observations"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Exercise instructions

Suppose you have a $Beta(1,1)$ prior distribution on the probability $\theta$ 
that a coin will yield a head when spun in a specified manner. The coin is 
independently spun ten times, and 'heads' appear fewer than 3 times. 
You are not told how many heads were seen, only that the number is 
less than 3. 

\begin{itemize}
\item[1)] Write down the equation for the exact posterior density for $\theta$. 
\item[2)] Discretize interval $[0,1]$ into 100 equally spaced intervals, calculate the posterior density at each of the discrete bins and draw the posterior density.
\item[3)] Using the discretized values of $\theta$ and their corresponding posterior density values a) draw the posterior cumulative density function of $\theta$ and b) calculate the posterior probability that $\theta<0.3$.
\item[4)] Draw the posterior density for $\theta$ in case where you are told that the exact number of heads is 1.  
What are theapparent differences between these two posterior densities?
\end{itemize}


## Model answer and grading

The posterior density function in case of censored observation is 

[FILL IN HERE WITH LATEX OR WRITE ON SEPARATE PAPER]

When discretizing  $\theta$ we can normalize the distribution as done below

```{r}
# vectorize th into 100 bins


# calculate the unnormalized density at each bin

# normalize the discretized probability densities

# calculate the cumulative distribution function

# plot the posterior density
par(mfrow=c(1,2))           # divide plot into 2 subplots
plot (....
# plot the posterior cumulative distribution function
plot (...

# calculate the probability that theta < 0.3


```

The posterior density funtion in case of $y=1$ observation is

[FILL IN HERE WITH LATEX OR WRITE ON SEPARATE PAPER]

```{r}
# calculate the density at each bin

# plot the unnormalized posterior
par(mfrow=c(1,2))           # divide plot into 2 subplots
plot (...
plot (...

```



*Grading* 3 points from correct answer to cases 1-3 and one point from correct answer to case 4.