---
title: 'Results: Beta-blockers exercise'
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Results: Beta-blockers exercise

Inference for a $2\times 2$ table: an experiment was performed to estimate 
the effect of beta-blockers on mortality of cardiac patients. A group of patients 
were randomly assigned to treatment and control groups: out of 674 patients 
receiving the control, 39 died, and out of 680 receiving the treatment, 22 died. 
Assume that the outcomes are independent and binomially distributed with 
probabilities of death $p_0$ and $p_1$ under the control and treatment, 
respectively. 

 * Set up a noninformative prior distribution for $(p_0,p_1)$ and draw samples from their posterior
 * Summarize the posterior distribution for the odds ratio $(p_1/(1-p_1))/(p_0(1-p_0))$ using the posterior mean, median and central 95\% credible interval.
 * Analyze and discuss the sensitivity of inference to the choice of the prior density.


```{r}

```
